Different rules for binocular combination of luminance flicker in cortical and subcortical pathways

### 2023-01-22 manuscriptNature.Rmd
Minor edits

How does the brain combine information across the eyes? In primary visual cortex, spatial patterns are subject to 'ocularity invariance', where monocular and binocular responses are rendered equal by mutual suppression between the eyes. Here we asked whether this invariance holds for pure changes in luminance. We collected steady-state EEG and pupillometry data simultaneously, and find strong deviations from ocularity invariance both in the cortex and also in the subcortical pathway that determines pupil diameter. In cortex, we find strong binocular facilitation, and negligible interocular suppression from cortical sources, whereas measurements of pupil diameter showed weaker facilitation and stronger suppression. This was not purely a consequence of the temporal stimulus parameters; pronounced binocular facilitation was also observed at faster flicker rates. Near-linear binocular combination was also found for the same stimuli using a perceptual matching task. A hierarchical Bayesian implementation of a standard binocular combination model confirms that interocular suppression was substantially weaker for our EEG and matching data compared with the pupillometry results. These findings illustrate that ocularity invariance is not a ubiquitous feature of visual processing, and that the brain can repurpose a generic algorithm for different functions by adjusting parameters such as the weight of suppression between pathways.

@@

The brain must combine information across multiple sensory inputs to derive a coherent percept of the external world. This involves a process of signal combination both within [@Baker2017] and between [@Ernst2002] the senses. Binocular vision is a useful test case for signal combination, as the inputs to the two eyes overlap substantially (in species with forward-facing eyes), and the neural locus is well-established [@Hubel1962]. Much of our knowledge about binocular combination derives from studies on the contrast response of the 'canonical' visual pathway, in which signals pass from the eyes to primary visual cortex (V1), via the lateral geniculate nucleus (LGN) [@Purves2008]. However, signals are also combined across the eyes in the network of subcortical nuclei that govern pupil diameter in response to absolute light levels [@McDougal2008] and much less is known about the computations that operate in these subcortical pathways. In addition, our primary purpose here is to investigate the computations governing signal combination in these two anatomically distinct pathways in response to luminance changes.

For pattern vision, binocular presentation confers greater sensitivity to low contrast targets than monocular presentation. This is known as binocular summation, with summation ratios (the relative improvement under binocular presentation) at detection threshold lying between $\sqrt{2}$ and 2 [@Campbell1965; @Baker2018]. This advantage is lost at high stimulus intensities, where both psychophysical performance (contrast discrimination thresholds) [@Legge1984; @Meese2006] and neural activity [@Moradi2009; @Baker2017] are approximately equal for monocular and binocular presentation. Contemporary models of binocular vision [@Ding2006; @Meese2006] advocate a process of interocular suppression that normalizes the two eyes inputs at high contrasts and negates the binocular advantage. This is consistent with our everyday experience of 'ocularity invariance' [@Baker2007]: perceived contrast does not change when one eye is opened and closed.

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink [@McDougal2008]. There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; @Wyatt1981). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus [@Angee2021; @Mathot2018; @McDougal2008; @Wang2015]. To account for the consensual response, these brain regions must  combine information from the left and right eyes [@Doesschate1967], yet the computation that achieves this is achieved is unclear.

@@

The pupillometry results are summarised in Fig. \@ref(fig:pupildata). The group average waveform for binocular presentation is shown in Fig. \@ref(fig:pupildata)a. There is a substantial pupil constriction at stimulus onset, followed by visible oscillations at the flicker frequency (2Hz, see waveform at foot). The average Fourier spectrum is displayed in Fig. \@ref(fig:pupildata)b, and shows a clear spike at 2 Hz, but no evidence of a second harmonic response at 4Hz. These results demonstrate that our paradigm can evoke measurable steady-state pupil responses at 2Hz.

@@

Fig. \@ref(fig:EEGdata) shows equivalent results, measured contemporaneously using EEG. Fig. \@ref(fig:EEGdata)a shows the group average waveform for binocular presentation, and Fig. \@ref(fig:EEGdata)b shows the Fourier spectrum for binocular presentation, both averaged across four posterior electrodes (_Oz_, _POz_, _O1_ and _O2_, marked on the inset scalp plots). Unlike for the pupillometry data, there are clear responses at both the first harmonic frequency (2Hz), and also the second harmonic frequency (4Hz). We therefore calculated contrast response functions at both first and second harmonic frequencies.

@@

When stimuli in both eyes flicker at 2Hz, the binocular responses at the first (Fig. \@ref(fig:EEGdata)c) and second (Fig. \@ref(fig:EEGdata)f) harmonics are substantially greater than the monocular responses, particularly at high contrasts. Analysis of variance on the complex values ($ANOVA^2_{circ}$) revealed a main effect of contrast (F(`r monbinEstats$dfA`,`r monbinEstats$dfR`) = `r round(monbinEstats$FratioA,digits=2)`, $p$ < 0.001) and an interaction effect (F(`r monbinEstats$dfAB`,`r monbinEstats$dfR`) = `r round(monbinEstats$FratioAB,digits=2)`, $p$ < 0.001), but no effect of condition ($p$ = `r round(monbinEstats$pvalB,digits=2)`) at the first harmonic, with a similar pattern of results obtained at the second harmonic. For the cross-frequency conditions (Fig. \@ref(fig:EEGdata)d,g), there was no appreciable effect of adding a 1.6Hz component on the response at 2Hz or 4Hz (no effect of condition, and no interaction). Similarly, there were no clear interocular interactions between frequencies in the responses at 1.6Hz (Fig. \@ref(fig:EEGdata)e) and 3.2Hz (Fig. \@ref(fig:EEGdata)h). This pattern of results suggests that processing of temporal luminance modulations happens in a more linear way in visual cortex (indexed by EEG), compared with subcortical pathways (indexed by pupillometry), and shows no evidence of interocular suppression.

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms [e.g. @Anstis1998; @Levelt1965; @Quaia2018]. We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

We were particularly interested in comparing the weight of interocular suppression across data sets. We therefore plot the posterior distributions for this parameter for all four data sets (see Fig. \@ref(fig:modelfigure)i). The key finding is that the pupillometry results (green distribution) display a much greater weight of interocular suppression compared with the other data sets (grey, purple and yellow distributions). There is no overlap between the pupillometry distribution and any of the other three. All four distributions are also meaningfully below a weight of 1 -- the value that previous work using grating stimuli would predict [@Meese2006; @Baker2017], and the peak location of our prior distribution (black curve). These results offer an explanation of the empirical data: the strong interocular suppression for the pupillometry data is consistent with the weak binocular facilitation, and measurable dichoptic masking observed using that method. The weaker suppression for the other experiments is consistent with the near-linear binocular facilitation effects, and absent dichoptic masking.

@@

The results of our main experiment were unexpected for both the pupillometry and the EEG measures. Previous studies investigating binocular combination of spatial patterns (i.e. sine wave grating stimuli) are all consistent with strong interocular suppression and weak binocular facilitation at high contrasts [@Baker2017; @Moradi2009; @Meese2006]. Our second experiment ruled out the possibility that these differences were due to the lower temporal frequency (2Hz) used here. However, there is evidence of more extensive binocular facilitation for a range of other stimuli. For example, @Quaia2018 observed a strong binocular facilitation (or 'supersummation') in the reflexive eye movement response to rapidly moving stimuli, and @Spitschan2019 report a similar result in archival data on melatonin suppression due to light exposure. In the auditory system, interaural suppression of amplitude modulation also appears to be weak when measured using a similar steady-state paradigm [@Baker2020]. Finally, psychophysical matching experiments using static stimuli also show near-linear behaviour for luminance increments [@Levelt1965; @Anstis1998; @Baker2012], though not for luminance decrements [@Anstis1998]. Overall, this suggests that strong interocular normalization may be specific to spatial pattern vision, and not a general feature of sensory signal combination.

Given the above, where does this leave our understanding of the overarching purpose of signal combination? @Baker2017 point out that strong suppression between channels that are subsequently summed is equivalent to a Kalman filter, which is the optimal method for combining two noisy inputs [see also @Ernst2002]. This account has intuitive appeal, and is consistent with other models that propose binocular combination as a means of redundancy reduction [@Li1994; @May2022]. One possibility is that optimal combination is useful for visual perception --- a critical system for interacting with the local environment --- and is therefore worth devoting the additional resource of inhibitory wiring between ocular channels. However the other examples given in the previous paragraph are primarily physiological responses (pupil size, eye movements, hormone release) that may benefit less from an increased signal-to-noise ratio, or otherwise be phylogenetically older than binocular pattern vision. Conceptualised another way, the brain can repurpose a generic architecture for different situational demands by adjusting parameter values (here the weight of interocular suppression) to achieve different outcomes. Our future work in this area intends to compare binocular combination for specific photoreceptor pathways, including different cone classes, and intrinsically photoreceptive retinal ganglion cells.

Pupil size determines the total amount of light falling on the retina. It is therefore the case that fluctuations in pupil diameter have a downstream effect on the signals reaching cortex. We did not incorporate such interactions into our computational model, though in principle this might be worthwhile. However we anticipate than any such effects would be small, since pupil modulations at 2Hz are in the order of 2% of overall diameter [e.g. @Spitschan2014]. It is also the case that cortical activity can modulate pupil diameter, usually through arousal and attention mechanisms [e.g. @Bradley2008]. We think it unlikely that these temporally coarse processes would have a differential effect on e.g. monocular and binocular stimulation conditions in our experiment, and any fluctuations during an experimental session (perhaps owing to fatigue) will be equivalent for our comparisons of interest. Therefore we make the simplifying assumption that the two pathways are effectively distinct, but hope to investigate this directly in future neuroimaging work.

Classic studies investigating the neurophysiological architecture of V1 reported that cells in cytochrome-oxidase 'blobs' [@Horton1981; @Livingstone1984] are biased towards low spatial frequencies [@Tootell1988; @Edwards1995], and relatively insensitive to stimulus orientation [@Horton1981; @Livingstone1984; though see @Economides2011]. As the blob regions are embedded within ocular dominance columns [@Horton1981], they are also largely monocular [@Livingstone1984; @Tychsen2004]. More recent work has reported psychophysical evidence for unoriented chromatic [@Gheiratmand2013] and achromatic [@Meese2011] mechanisms, that also appear to be monocular. Our use of luminance flicker might preferentially stimulate these mechanisms, perhaps explaining why our EEG data show little evidence of binocular interactions. Indeed, our EEG results could potentially be explained by a model involving entirely non-interacting monocular channels, with the binocular facilitation effects we find (e.g. Fig. \@ref(fig:BSratios) & \@ref(fig:TFdata)) owing to additivity of the electrophysiological response across independent monocular cells, rather than requiring binocular neurons. However our matching data (Fig. \@ref(fig:matchingdata)), as well as everyday perceptual experience, indicates that luminance signals must be combined physiologically at some stage, and that this process still involves weak or absent interocular suppression.

@@

Thirty (20 females), twelve (7 females) and ten (3 females) adult participants, whose ages ranged from 18 to 45, were recruited for Experiments 1, 2 and 3 respectively. All participants had normal or corrected to normal binocular vision, and gave written informed consent in accordance to the Declaration of Helsinki. Our procedures were approved by the Ethics Committee of the Department of Psychology at the University of York (identification number 792).

@@

The stimuli were two discs of achromatic flickering light with a diameter of 3.74 degrees, presented on a black background. The same stimuli were used for all three experiments. Four dark red lines were added around both discs to help with their fusion into one binocular disc (see insert in Fig. \@ref(fig:pupildata)b for an example of the fused stimulus). The discs were viewed through a four-mirror stereoscope, which used front silvered mirrors to avoid internal reflections, and meant that participants saw a single fused disc. The use of a stereoscope allowed us to modulate the stimuli in three different ocular configurations: monocular, binocular, and dichoptic. Note that during monocular presentation of flicker, the unstimulated eye still saw the static (non-flickering) disc of mean luminance.

All stimuli had a mean luminance of 42 cd/m$^2$ on an Iiyama VisionMaster™ Pro 510 display (800 x 600 pixels, 60 Hz refresh rate), which was gamma corrected using a Minolta LS-110 photometer (Minolta Camera Co. Ltd., Japan). For experiments 1 and 2, the stimuli were presented using Psychopy (v3.0.7). For experiment 3, the stimuli were presented using Psychopy (v2022.1.1).

EEG data were collected for Experiments 1 and 2 using a 64-electrode ANT WaveGuard cap and the signals were recorded at 1 kHz using the ASA software (ANT Neuro, Netherlands). Pupillometry data were collected for Experiment 1 using a binocular Pupil Core eye-tracker (Pupil Labs GmbH, Berlin, Germany; @Kassner2014) running at 120 Hz, and the signals were recorded with the Pupil Capture software.

@@

We included six distinct ocular conditions, each at five temporal contrast levels (combined factorially) relative to the mean luminance: 6, 12, 24, 48 and 96\%. Contrast was defined as temporal Michelson contrast; the difference between maximum and minimum luminances, scaled by the mean and expressed as a percentage. In the first three conditions, the discs flickered at 2 Hz, in either a monocular, binocular, or dichoptic arrangement. In the dichoptic condition the non-target eye saw a fixed contrast of 48\%. In the remaining three conditions (termed the cross-frequency conditions) one eye's disc flickered at 1.6Hz, and the other eye's disc flickered at 2Hz. We also tested monocular (one eye sees 1.6Hz flicker, the other sees mean luminance), binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 2Hz, mask contrast of 48\% at 1.6Hz in the other eye) arrangements. The rationale for flickering both eyes at 2Hz is that we can then measure summation behaviour between the eyes in the pupil and EEG response at 2Hz. The rationale for flickering the eyes at different frequencies is that this permits measurement of suppression between the eyes (i.e. the reduction in the 2Hz response when a 1.6Hz mask component is added to the other eye). We counterbalanced presentation of the target stimulus across the left and right eyes.

@@

EEG data were converted from the ANT-EEProbe format to a compressed csv text file using a custom Matlab script and components of the EEGlab toolbox [@Delorme2004]. The data for each participant were then loaded into R for analysis, where a ten-second waveform for each trial at each electrode was extracted (omitting the first two seconds). The Fourier transform of each waveform was calculated, and the complex spectrum stored in a matrix. All repetitions of each condition were then averaged for each electrode. They were then averaged across four occipital electrodes (\emph{POz}, \emph{Oz}, \emph{O1}, \emph{O2}), to obtain individual results. Finally, these were averaged across participants to obtain the group results. All averaging was performed in the complex domain and therefore retained the phase information (i.e. coherent averaging), and at each stage we excluded data points with a Mahalanobis distance exceeding $D$ = `r SDthresh` from the complex-valued mean [see @Baker2021]. For statistical comparisons of complex-valued data, we use the $ANOVA^2_{circ}$ statistic described by @Baker2021. This is a multivariate extension of ANOVA that assumes equal variance of the real and imaginary Fourier components, or equivalently, an extension of the $T^2_{circ}$ statistic of @Victor1991 that can compare more than two conditions.

A similar analysis pipeline was adopted for the pupillometry data. The data were converted from mp4 videos to a csv text file using the Pupil Player software [@Kassner2014], which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where again a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. Again, we used coherent averaging, and excluded outlying data points in the same way as for the EEG data. Note that previous pupillometry studies using luminance flicker have tended to fit a single sine-wave at the fundamental frequency, rather than using Fourier analysis [e.g. @Spitschan2014]. The Fourier approach is more robust to noise at other frequencies and has been used in some previous studies [see @Barrionuevo2014; @Barrionuevo2016]. Additionally, it makes the pupillometry analysis consistent with standard practice in steady-state EEG analysis [e.g. @Figueira2022].

@@

To describe our data, we chose a model of binocular contrast gain control with the same general form as the first stage of the model proposed by @Meese2006. The second gain control stage was omitted [consistent with @Baker2017] to simplify the model and reduce the number of free parameters. The response of the left eye's channel is given by:

@@

Despite being derived from the model proposed by @Meese2006, the simplifications applied to this architecture make it very similar to other models [e.g. @Ding2006; @Legge1984; @Schrodinger1926; @Doesschate1967]. In particular we fixed the numerator exponent at 2 in our model, because otherwise this value tends to trade off with the weight of interocular suppression [see @Baker2012; @Kingdom2015]. Our key parameter of interest is the weight of interocular suppression. Large values around $w$ = 1 result in a very small or nonexistent binocular advantage at suprathreshold contrasts, consistent with previous work using grating stimuli [@Baker2017]. Low values around $w$ = 0 produce substantial, near-linear binocular facilitation [@Baker2020].

>>>>

How does the human brain combine information across the eyes? It has been known for many years that cortical normalisation mechanisms implement ‘ocularity invariance’: equalising neural responses to spatial patterns presented either monocularly or binocularly. Here we used a novel combination of electrophysiology, psychophysics, pupillometry and computational modelling to ask whether this invariance also holds for flickering luminance stimuli with no spatial contrast. We find dramatic violations of ocularity invariance for these stimuli, both in cortex and also in the subcortical pathways that govern pupil diameter. Specifically, we find substantial binocular facilitation in both pathways with the effect being strongest in cortex. Near-linear binocular additivity (instead of ocularity invariance) was also found using a perceptual luminance matching task. Ocularity invariance is therefore not a ubiquitous feature of visual processing, and the brain appears to repurpose a generic normalisation algorithm for different visual functions by adjusting the amount of interocular suppression.

@@

The brain must combine information across multiple sensory inputs to derive a coherent percept of the external world. This involves a process of signal combination both within [@Baker2017] and between [@Ernst2002] the senses. Binocular vision is a useful test case for signal combination, as the inputs to the two eyes overlap substantially (in species with forward-facing eyes), and the neural locus is well-established [@Hubel1962]. Much of our knowledge about binocular combination derives from studies on the contrast response of the 'canonical' visual pathway, in which signals pass from the eyes to primary visual cortex (V1), via the lateral geniculate nucleus (LGN) [@Purves2008]. However, signals are also combined across the eyes in the network of subcortical nuclei that govern pupil diameter in response to absolute light levels [@McDougal2008] and much less is known about the computations that operate in these subcortical pathways. Our primary purpose here is to investigate the computations governing signal combination in these two anatomically distinct pathways in response to luminance changes.

For pattern vision, binocular presentation confers greater sensitivity to low contrast targets than monocular presentation. This is known as binocular summation[@Campbell1965; @Baker2018], with summation ratios (the relative improvement under binocular presentation) at detection threshold lying between $\sqrt{2}$ and 2. This advantage is lost at high stimulus intensities, where both psychophysical performance (contrast discrimination thresholds) [@Legge1984; @Meese2006] and neural activity [@Moradi2009; @Baker2017] are approximately equal for monocular and binocular presentation. Contemporary models of binocular vision [@Ding2006; @Meese2006] advocate a process of interocular suppression that normalizes the two eyes inputs at high contrasts and negates the binocular advantage. This is consistent with our everyday experience of 'ocularity invariance' [@Baker2007]: perceived contrast does not change when one eye is opened and closed.

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink [@McDougal2008]. There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; @Wyatt1981). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus [@Angee2021; @Mathot2018; @McDougal2008; @Wang2015]. To account for the consensual response, these brain regions must  combine information from the left and right eyes [@Doesschate1967], yet the computation that achieves this is unclear.

@@

The pupillometry results for our first experiment are summarised in Fig. \@ref(fig:pupildata). The group average waveform for binocular presentation is shown in Fig. \@ref(fig:pupildata)a. There is a substantial pupil constriction at stimulus onset, followed by visible oscillations at the flicker frequency (2Hz, see waveform at foot). The average Fourier spectrum is displayed in Fig. \@ref(fig:pupildata)b, and shows a clear spike at 2 Hz, but no evidence of a second harmonic response at 4Hz. These results demonstrate that our paradigm can evoke measurable steady-state pupil responses at 2Hz.

@@

Fig. \@ref(fig:EEGdata) shows equivalent results, measured contemporaneously using EEG. Fig. \@ref(fig:EEGdata)a shows the group average waveform for binocular presentation, and Fig. \@ref(fig:EEGdata)b shows the Fourier spectrum for binocular presentation, both averaged across four posterior electrodes (_Oz_, _POz_, _O1_ and _O2_, marked on the inset scalp plots). Unlike for the pupillometry data, there are clear responses at both the first harmonic frequency (2Hz), and also the second harmonic frequency (4Hz). We therefore derived contrast response functions at both first and second harmonic frequencies.

@@

When stimuli in both eyes flicker at 2Hz, the binocular responses at the first (Fig. \@ref(fig:EEGdata)c) and second (Fig. \@ref(fig:EEGdata)f) harmonics are substantially greater than the monocular responses, particularly at high contrasts. Analysis of variance on the complex values ($ANOVA^2_{circ}$) revealed a main effect of contrast (F(`r monbinEstats$dfA`,`r monbinEstats$dfR`) = `r round(monbinEstats$FratioA,digits=2)`, $p$ < 0.001) and an interaction effect (F(`r monbinEstats$dfAB`,`r monbinEstats$dfR`) = `r round(monbinEstats$FratioAB,digits=2)`, $p$ < 0.001), but no effect of condition ($p$ = `r round(monbinEstats$pvalB,digits=2)`) at the first harmonic, with a similar pattern of results obtained at the second harmonic (note that the absence of a main effect of condition is expected, given that there should be negligible differences between conditions at low contrasts, and any differences are therefore captured by the interaction term). For the cross-frequency conditions (Fig. \@ref(fig:EEGdata)d,g), there was no appreciable effect of adding a 1.6Hz component on the response at 2Hz or 4Hz (no effect of condition, and no interaction). Similarly, there were no clear interocular interactions between frequencies in the responses at 1.6Hz (Fig. \@ref(fig:EEGdata)e) and 3.2Hz (Fig. \@ref(fig:EEGdata)h). This pattern of results suggests that processing of temporal luminance modulations happens in a more linear way in visual cortex (indexed by EEG), compared with subcortical pathways (indexed by pupillometry), and shows no evidence of interocular suppression.

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms [@Anstis1998; @Levelt1965; @Quaia2018]. We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

We were particularly interested in comparing the weight of interocular suppression across data sets. We therefore plot the posterior distributions for this parameter for all four data sets (see Fig. \@ref(fig:modelfigure)i). The key finding is that the pupillometry results (green distribution) display a much greater weight of interocular suppression compared with the other data sets (grey, purple and yellow distributions). There is no overlap between the pupillometry distribution and any of the other three. All four distributions are also meaningfully below a weight of 1 -- the value that previous work using grating stimuli would predict [@Meese2006; @Baker2017], and the peak location of our prior distribution (black curve). These results offer an explanation of the empirical data: the strong interocular suppression for the pupillometry data is consistent with the weak binocular facilitation, and significant dichoptic masking observed using that method. The weaker suppression for the other experiments is consistent with the near-linear binocular facilitation effects, and absent dichoptic masking.

@@

The results of our main experiment were unexpected for both the pupillometry and the EEG measures. Previous studies investigating binocular combination of spatial patterns (i.e. sine wave grating stimuli) are universally consistent with strong interocular suppression and weak binocular facilitation at high contrasts [@Baker2017; @Moradi2009; @Meese2006]. Our second experiment ruled out the possibility that these differences were due to the lower temporal frequency (2Hz) used here. However, there is evidence of more extensive binocular facilitation for a range of other stimuli. For example, Quaia et al.@Quaia2018 observed a strong binocular facilitation (or 'supersummation') in the reflexive eye movement response to rapidly moving stimuli, and Spitschan and Cajochen@Spitschan2019 report a similar result in archival data on melatonin suppression due to light exposure. In the auditory system, interaural suppression of amplitude modulation also appears to be weak when measured using a similar steady-state paradigm [@Baker2020]. Finally, psychophysical matching experiments using static stimuli also show near-linear behaviour for luminance increments [@Levelt1965; @Anstis1998; @Baker2012], though not for luminance decrements [@Anstis1998]. Overall, this suggests that strong interocular normalization may be specific to spatial pattern vision, and not a general feature of binocular signal combination.

Given the above, where does this leave our understanding of the overarching purpose of signal combination? We previously@Baker2017 pointed out that a model with strong suppression between channels that are subsequently summed is equivalent to a Kalman filter, which is the optimal method for combining two noisy inputs (see also [@Ernst2002]). This account has intuitive appeal, and is consistent with other models that propose binocular combination as a means of redundancy reduction [@Li1994; @May2022]. One possibility is that optimal combination is useful for visual perception --- a critical system for interacting with the local environment --- and is therefore worth devoting the additional resource of inhibitory wiring between ocular channels. However the other examples given in the previous paragraph are primarily physiological responses (pupil size, eye movements, hormone release) that may benefit less from an increased signal-to-noise ratio, or are otherwise phylogenetically older than binocular pattern vision. Conceptualised another way, the brain can repurpose a generic architecture for different situational demands by adjusting parameter values (here the weight of interocular suppression) to achieve different outcomes. Our future work in this area intends to compare binocular combination for specific photoreceptor pathways, including different cone classes, and intrinsically photoreceptive retinal ganglion cells.

Pupil size determines the total amount of light falling on the retina. It is therefore the case that fluctuations in pupil diameter have a downstream effect on the signals reaching cortex. We did not incorporate such interactions into our computational model, though in principle this might be worthwhile. However we anticipate than any such effects would be small, since pupil modulations at 2Hz are in the order of 2% of overall diameter [@Spitschan2014]. It is also the case that cortical activity can modulate pupil diameter, usually through arousal and attention mechanisms [@Bradley2008]. We think it unlikely that these temporally coarse processes would have a differential effect on e.g. monocular and binocular stimulation conditions in our experiment, and any fluctuations during an experimental session (perhaps owing to fatigue) will be equivalent for our comparisons of interest. Therefore we make the simplifying assumption that the two pathways are effectively distinct, but intend to investigate this directly in future neuroimaging work.

Classic studies investigating the neurophysiological architecture of V1 reported that cells in cytochrome-oxidase 'blobs' [@Horton1981; @Livingstone1984] are biased towards low spatial frequencies [@Tootell1988; @Edwards1995], and relatively insensitive to stimulus orientation [@Horton1981; @Livingstone1984] (though see @Economides2011). As the blob regions are embedded within ocular dominance columns [@Horton1981], they are also largely monocular [@Livingstone1984; @Tychsen2004]. More recent work has reported psychophysical evidence for unoriented chromatic [@Gheiratmand2013] and achromatic [@Meese2011] mechanisms, that also appear to be monocular. Our use of luminance flicker might preferentially stimulate these mechanisms, perhaps explaining why our EEG data show little evidence of binocular interactions. Indeed, our EEG results could potentially be explained by a model involving entirely non-interacting monocular channels, with the binocular facilitation effects we find (e.g. Fig. \@ref(fig:BSratios) & \@ref(fig:TFdata)) owing to additivity of the electrophysiological response across independent monocular cells, rather than requiring binocular neurons. However our matching data (Fig. \@ref(fig:matchingdata)), as well as everyday perceptual experience, indicates that luminance signals must be combined physiologically at some stage, and that this process still involves weak or absent interocular suppression.

@@

Thirty (20 F, 10 M), twelve (7 F, 5 M) and ten (3 F, 7 M) adult participants, whose ages ranged from 18 to 45, were recruited for Experiments 1, 2 and 3 respectively. All participants had normal or corrected-to-normal binocular vision, and gave written informed consent in accordance with the Declaration of Helsinki. Our procedures were approved by the Ethics Committee of the Department of Psychology at the University of York (identification number 792).

@@

The stimuli were two monocularly presented discs of achromatic flickering light with a diameter of 3.74 degrees, displayed against a black background. The same stimuli were used for all three experiments. Four dark red lines were added around each monocular disc to help with their fusion into a single binocular disc (see insert in Fig. \@ref(fig:pupildata)b for an example of the fused stimulus). The discs were viewed through a four-mirror stereoscope, which used front silvered mirrors to avoid internal reflections, and meant that participants perceived a single fused disc. The use of a stereoscope allowed us to modulate the stimuli in three different ocular configurations: monocular, binocular, and dichoptic. Note that during monocular presentation of flicker, the unstimulated eye still saw the static (non-flickering) disc of mean luminance.

All stimuli had a mean luminance of 42 cd/m$^2$, and were presented on an Iiyama VisionMaster™ Pro 510 display (800 x 600 pixels, 60 Hz refresh rate), which was gamma corrected using a Minolta LS-110 photometer (Minolta Camera Co. Ltd., Japan). All stimuli were presented using Psychopy [@Peirce2019] (v3.0.7 for Experiments 1 and 2; v2022.1.1 for Experiment 3).

EEG data were collected for Experiments 1 and 2 using a 64 electrode ANT WaveGuard cap and the signals were recorded at 1 kHz using the ASA software (ANT Neuro, Netherlands). Pupillometry data were collected for Experiment 1 using a binocular Pupil Core eye-tracker (Pupil Labs GmbH, Berlin, Germany @Kassner2014) running at 120 Hz, and the signals were recorded with the Pupil Capture software.

@@

We included six distinct ocular conditions, each at five temporal contrast levels (combined factorially) relative to the mean luminance: 6, 12, 24, 48 and 96\%. Contrast was defined as temporal Michelson contrast; the difference between maximum and minimum luminances, scaled by the mean and expressed as a percentage. In the first three conditions, the discs flickered at 2 Hz, in either a monocular, binocular, or dichoptic arrangement. In the dichoptic condition the non-target eye saw a fixed contrast of 48\%. In the remaining three conditions (termed the cross-frequency conditions) one eye's disc flickered at 2Hz, and the other eye's disc flickered at 1.6Hz. We also tested monocular (one eye sees 1.6Hz flicker, the other sees mean luminance), binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 2Hz, mask contrast of 48\% at 1.6Hz in the other eye) arrangements. The rationale for flickering both eyes at 2Hz is that we can measure summation behaviour between the eyes in the pupil and EEG response at 2Hz. The rationale for flickering the eyes at different frequencies is that this permits measurement of suppression between the eyes (i.e. the reduction in the 2Hz response when a 1.6Hz mask component is added to the other eye). We counterbalanced presentation of the target stimulus across the left and right eyes.

@@

EEG data were converted from the ANT-EEProbe format to a compressed csv text file using a custom Matlab script and components of the EEGlab toolbox [@Delorme2004]. The data for each participant were then loaded into R for analysis, where a ten-second waveform for each trial at each electrode was extracted (omitting the first two seconds). The Fourier transform of each waveform was calculated, and the complex spectrum stored in a matrix. All repetitions of each condition were then averaged for each electrode. They were then averaged across four occipital electrodes (\emph{POz}, \emph{Oz}, \emph{O1}, \emph{O2}), to obtain individual results. Finally, these were averaged across participants to obtain the group results. All averaging was performed in the complex domain and therefore retained the phase information (i.e. coherent averaging), and at each stage we excluded data points with a Mahalanobis distance exceeding $D$ = `r SDthresh` from the complex-valued mean [@Baker2021]. For statistical comparisons of complex-valued data, we use the $ANOVA^2_{circ}$ statistic described previously @Baker2021. This is a multivariate extension of ANOVA that assumes equal variance of the real and imaginary Fourier components, or equivalently, an extension of the $T^2_{circ}$ statistic @Victor1991 that can compare more than two conditions.

A similar analysis pipeline was adopted for the pupillometry data. The data were converted from mp4 videos to a csv text file using the Pupil Player software [@Kassner2014], which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where again a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. Again, we used coherent averaging, and excluded outlying data points in the same way as for the EEG data. Note that previous pupillometry studies using luminance flicker have tended to fit a single sine-wave at the fundamental frequency, rather than using Fourier analysis @Spitschan2014. The Fourier approach is more robust to noise at other frequencies and has been used in some previous studies [@Barrionuevo2014; @Barrionuevo2016]. Additionally, it makes the pupillometry analysis consistent with standard practice in steady-state EEG analysis [@Figueira2022].

@@

To describe our data, we chose a model of binocular contrast gain control with the same general form as the first stage of the model proposed by Meese et al. [@Meese2006]. The second gain control stage was omitted, consistent with previous work [@Baker2017], to simplify the model and reduce the number of free parameters. The response of the left eye's channel is given by:

@@

Despite being derived from the model proposed by Meese et al. [@Meese2006], the simplifications applied to this architecture make it very similar to other models [@Ding2006; @Legge1984; @Schrodinger1926; @Doesschate1967]. In particular we fixed the numerator exponent at 2 in our model, because otherwise this value tends to trade off with the weight of interocular suppression [@Baker2012; @Kingdom2015]. Our key parameter of interest is the weight of interocular suppression. Large values around $w$ = 1 result in a very small or nonexistent binocular advantage at suprathreshold contrasts, consistent with previous work using grating stimuli [@Baker2017]. Low values around $w$ = 0 produce substantial, near-linear binocular facilitation [@Baker2020].
